---
title: "NYDP Shooting Incident Data"
author: "Khoa Bui"
output:
  pdf_document: default
  html_document: default
date: "03/05/2024"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dataset Description

This is a breakdown of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year.

This data is manually extracted every quarter and reviewed by the Office of Management Analysis and Planning before being posted on the NYPD website. Each record represents a shooting incident in NYC and includes information about the event, the location and time of occurrence. In addition, information related to suspect and victim demographics is also included. This data can be used by the public to explore the nature of shooting/criminal activity. Please refer to the attached data footnotes for additional information about this dataset.

Source: [NYPD Shooting Incident](httpds://catalog.data.gov/dataset/nypd-shooting-incident-data-historic)

## Step 0: Install And/Or Import Libraries

(Optional): If you don't have any of these packages installed yet, uncomment these lines below and run it

(Required): I used extra library called "dplyr", so please make sure you at least install that packages before knit my file.

```{r library, message=FALSE, warning=FALSE}
# install.packages("tidyverse")
# install.packages("lubridate")
# install.packages("ggplot2")
# install.packages("dplyr")
library(dplyr)
library(tidyverse)
library(lubridate)
library(ggplot2)
```

## Step 1: Import Dataset

```{r load}
dataUrl <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

# read_cvs is used because it being used to read comma seperated values file.
shootingData = read_csv(dataUrl)
```

## Step 1.5: How to read dataset

Eliminated some column because didn't need to use. The definition of each in-use column is included but please feel free to explore the rest using the NYPD Shooting Incident website provided above, if needed.

### Row Description

- Each __row__ in this dataset is presenting unique __shooting incident__.

### Column Description

- __INCIDENT_KEY__: (dbl) Unique incident ID assigned for each incident
- __OCCUR_DATE__: (chr) Date of shooting incident in mm/dd/yyyy
- __OCCUR_TIME__: (time) Time of the shooting incident in hh/mm/ss using 24hours system
- __BORO__: (chr) Borough where the shooting incident occurred
- __STATISTICAL_MURDER_FLAG__: (lgl) True/ False if the system became a murder
- __PERP_AGE_GROUP__: (chr) Perpetrator's age group
- __PERP_SEX__: (chr) Perpetrator's sex identification
- __PERP_RACE__: (chr) Perpetrator's race identification
- __VIC_AGE_GROUP__: (chr) Victim's age group
- __VIC_SEX__: (chr) Victim's sex identification
- __VIC_RACE__: (chr) Victim's race

## Step 2: Tidy and Transform Data

Eliminating these column from the dataset: LOC_OF_OCCUR_DESC, PRECINCT, JURISDICTION_CODE, LOC_CLASSFCTN_DESC, LOCATION_DESC, X_COORD_CD, Y_COORD_CD, Latitude, Longitude, Lon_Lat.

``` {r remove_unused}
# Remove unused column from dataset
shootingData <- shootingData %>% select(-c(
                      LOC_OF_OCCUR_DESC,
                      PRECINCT,
                      JURISDICTION_CODE,
                      LOC_CLASSFCTN_DESC,
                      LOCATION_DESC,
                      X_COORD_CD,
                      Y_COORD_CD,
                      Latitude,
                      Longitude,
                      Lon_Lat
                      ))
glimpse(shootingData) # Print dataset after removed unused column
```

Since __OCCUR_DATE__: (Char) Date of shooting incident in mm/dd/yyyy, we should convert it to data data type instead

``` {r convert_occur_date}
shootingData$OCCUR_DATE <- mdy(shootingData$OCCUR_DATE)
```

We know that __PERP_AGE_GROUP__ and __VIC_AGE_GROUP__ is having data type of "chr", but we also want to remove all data that doesn't make senes, so we have to manually cross checking with the actual excel file to remove it for now

``` {r remove_error_values}
# Remove error values in dataset
shootingData = subset(shootingData, PERP_AGE_GROUP!="1020" & PERP_AGE_GROUP!="224" & PERP_AGE_GROUP!="940" & VIC_AGE_GROUP!="1022")
```

### Key observations on data type conversion are:

- __OCCUR_DATE__: Will be used to get Year.
- __BORO__: Will be treated as a factor.
- __PERP_AGE_GROUP__: Will be treated as a factor.
- __PERP_SEX__: Will be treated as a factor.
- __PERP_RACE__: Will be treated as a factor.
- __VIC_AGE_GROUP__: Will be treated as a factor.
- __VIC_SEX__: Will be treated as a factor.
- __VIC_RACE__: Will be treated as a factor.
- __STATISTICAL_MURDER_FLAG__: Will be treated as a factor

### Unknown/ Missing Value

The reason behind there is missing value in __PERP_AGE_GROUP__, __PERP_SEX__, __PERP_RACE__ is because these case is till a cold case/ unsolved case.
In other to ensure our data is correct, I will avoid using these column, by avoid doing study on PERP AGE, SEX, and RACE since to me, it is not enough data to study.
Instead, I will study how many unsolved case, how long the case have been unsolved for, and all the related things.
To do so, I will treat all "NA" or "UNKNOWN" as "Unknown"

``` {r tidy_factor_transform_data}
# Tidy and transform data
shootingData = shootingData %>% 
  replace_na(list(PERP_AGE_GROUP = "Unknown", PERP_SEX = "Unknown", PERP_RACE = "Unknown"))
shootingData <- shootingData %>%
  mutate_at(c("PERP_AGE_GROUP","PERP_SEX","PERP_RACE"), list(~ifelse(. == "(null)", "Unknown", .)))
shootingData <- shootingData %>%
  mutate_at(c("PERP_AGE_GROUP","PERP_RACE","VIC_AGE_GROUP",
              "VIC_RACE"), list(~ifelse(. == "UNKNOWN", "Unknown", .)))
shootingData <- shootingData %>%
  mutate_at(c("PERP_SEX","VIC_SEX"), list(~ifelse(. == "U", "Unknown", .)))

# Add OCCUR_YEAR to the dataset
shootingData$OCCUR_YEAR <- year(shootingData$OCCUR_DATE)

# Transform to factor
shootingData$BORO = as.factor(shootingData$BORO)
shootingData$PERP_AGE_GROUP = as.factor(shootingData$PERP_AGE_GROUP)
shootingData$PERP_SEX = as.factor(shootingData$PERP_SEX)
shootingData$PERP_RACE = as.factor(shootingData$PERP_RACE)
shootingData$VIC_AGE_GROUP = as.factor(shootingData$VIC_AGE_GROUP)
shootingData$VIC_SEX = as.factor(shootingData$VIC_SEX)
shootingData$VIC_RACE = as.factor(shootingData$VIC_RACE)
shootingData$STATISTICAL_MURDER_FLAG <- factor(shootingData$STATISTICAL_MURDER_FLAG)
```

Summary Statistic

``` {r summary}
summary(shootingData)
```

## Step 3: Visualizations and Analysis

###### 1. Non-murder by year and murders by year 

```{r incident_vs_murder}
shootingData %>%
  filter(STATISTICAL_MURDER_FLAG == FALSE) %>%
  ggplot(aes(x = OCCUR_YEAR)) +
  geom_bar() +
  labs(title = "Non-murder per year",
       x = "Years",
       y = "Non-murders") +
  theme_minimal()

shootingData %>%
  filter(STATISTICAL_MURDER_FLAG == TRUE) %>%
  ggplot(aes(x = OCCUR_YEAR)) +
  geom_bar() +
  labs(title = "Murders Per Year",
       x = "Years",
       y = "Murders") +
  theme_minimal()

table(shootingData$OCCUR_YEAR, shootingData$STATISTICAL_MURDER_FLAG)
```

Based on graph and data table above, since murder is way less than non-murder, the rate of change is a bit different, but in the safe zone to say it is align with each other.

#### 2. Solved vs Unsolved Case by Year

Unsolved Case

```{r unsolved}
shootingData %>%
  filter(PERP_AGE_GROUP == "Unknown") %>%
  ggplot(aes(x = OCCUR_YEAR)) +
  geom_bar() +
  labs(title = "Unsolved Case Per Year",
       x = "Years",
       y = "Unsolved") +
  theme_minimal()
```

``` {r solved_case}
shootingData %>%
  filter(PERP_AGE_GROUP != "Unknown") %>%
  ggplot(aes(x = OCCUR_YEAR)) +
  geom_bar() +
  labs(title = "Solved Case Per Year",
       x = "Years",
       y = "Solved") +
  theme_minimal()

# I used PERP_AGE_GROUP because if the value is Unknown
# that mean the case is not solved/ missing data
table(shootingData$OCCUR_YEAR, shootingData$PERP_AGE_GROUP)
```

Based on these data above, is it safe to say that, the further the year, the more case didn't solved, isn't it? Maybe. But I also have another theory. Maybe the data for these years was before when we actually collect data, so these data wasn't collected properly. This is something we need to dive deep to ensure we known our dataset inside out.

#### 3. Building logistic regression model to predict if the victim will be survied

Logistic regression is an instance of classification technique that you can use to predict a qualitative response. I will use logistic regression models to estimate the probability that a murder case belongs to a particular victim's profile.

The output shows the coefficients, their standard errors, the z-values, and the associated p-values.The logistic regression coefficients give the change in the log odds of the outcome for a one unit increase in the predictor variable.

```{r}
# Logistics Regression 
glm.fit <- glm(STATISTICAL_MURDER_FLAG ~ VIC_AGE_GROUP + VIC_RACE + VIC_SEX, 
               data = shootingData, family = binomial)
summary(glm.fit)
```

__Notable Findings__: The age bracket of the __affected individual__ appears pivotal in predicting their chances of surviving a shooting incident. Notably, individuals within the __< 18__ and __18-24__ age categories exhibit the highest likelihood of surviving such incidents. Conversely, the probability of survival steadily __diminishes__ across successive age groups. Particularly grim are the outcomes for shootings involving individuals __aged 65 and above__, where fatalities are prevalent.


### Area to dive deep

- Are there additional variables besides age that could serve as indicators for the fatality of a shooting incident, such as the victim's sex?

## Step 4: Conclusion && Identify Bias

### Conclusion

We explored how many unsolved case, among with if the shot is fatal what is the varibles to make that shot fatal.

### Room for bias

One of the most common bias is that if the victim is in one race, normally the perp will also be in the same Race. There is also another bias that if there is a gun shot, it should be fatal. For the scope of this project, I didn't touch perp race for that reason, but dive deep into if the shot fatal or not.